---
title: "Analysis of New York Shooting Incident"
author: "Laurence Tan"
date: "`r Sys.Date()`"
output:
  word_document: default
  pdf_document: default
  html_document: default
---

## Introduction
There has been a concern in New York City (NYC) regarding shootings related to criminal activity. Would increasing police presence reduce these shooting incidents? Intuitively, one could argue that more precincts could deter criminal activities and potentially lower shooting incidents. However, alternative solutions, such as mental health resources and social services, could address the root causes of violence more effectively. This analysis aims to examine the data from NYPD shooting incidents, correlating the number of precincts (indicating police presence) with the number of shooting incidents. This study seeks to inform evidence-based policies that can promote a safer NYC.

## Data Import and Initial Exploration

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
## Load the require libraries
library(tidyverse)


```

### Importing Data
Each record from the NYPD Shooting Incident Data (Historic) starting from 2006 in NYC includes the precinct, location, time of occurrence, age group, sex, and race of both perpetrators and victims. The CSV file can be downloaded from https://catalog.data.gov/dataset/nypd-shooting-incident-data-year-to-date
```{r load_dataset}
# Importing the data from CSV
# identifier of the data can be found in https://data.cityofnewyork.us/api/views/833y-fsy8
csv_file <- "https://data.cityofnewyork.us/api/views/833y-fsy8/rows.csv"
NY_shooting_cases = read_csv(csv_file)

```

### Examine Data Structure
```{r check_dataset}
# Examine the first few rows of the dataset
head(NY_shooting_cases)

# Examine the data type
str(NY_shooting_cases)

# Summerize the dataset
summary(NY_shooting_cases)


```

## Data Cleaning
### Replace Null Values
```{r find_null_values}
# Find the null values
null_values <- colSums(is.na(NY_shooting_cases))
null_values

# Replace null values with "UNKNOWN" so not to disrupt calculations or 
# aggregations. This also provides a clear indicator that data 
# for certain fields were not recorded.
# Geographic coordinates (Latitude and Longitude) better to leave them as NA.
# Leaving JURISDICTION_CODE as NA as not in use for this analysis.
NY_shooting_cases_Updated <- NY_shooting_cases %>%
  mutate(
    PERP_AGE_GROUP = replace_na(PERP_AGE_GROUP, "UNKNOWN"),
    PERP_SEX = replace_na(PERP_SEX, "UNKNOWN"),
    PERP_RACE = replace_na(PERP_RACE, "UNKNOWN"),
    LOC_CLASSFCTN_DESC = replace_na(LOC_CLASSFCTN_DESC, "UNKNOWN"),
    LOCATION_DESC  = replace_na(LOCATION_DESC , "UNKNOWN"),    
    LOC_OF_OCCUR_DESC  = replace_na(LOC_OF_OCCUR_DESC , "UNKNOWN")
  )

#Verify null values updated
null_values <- colSums(is.na(NY_shooting_cases_Updated))
null_values

```

### Convert Data Types
```{r Convert_OCCUR_DATE_Date_Format}
# Convert OCCUR_DATE to Date format to YYYY-MM-DD
# Convert to time format then extract the hour)
NY_shooting_cases_Updated <- NY_shooting_cases_Updated %>%
  mutate(OCCUR_DATE = mdy(OCCUR_DATE),
         OCCUR_MONTH = month(OCCUR_DATE),
         OCCUR_YEAR = year(OCCUR_DATE),
         OCCUR_TIME = hms::as_hms(OCCUR_TIME),
         OCCUR_HOUR = hour(OCCUR_TIME)
         )

```


## Exploratory Data Analysis
### Incidents by Location
Notably, Brooklyn and the Bronx have significantly higher shooting incidents compared to other locations. Further analysis should be carried out to understand the underlying reasons for the high shooting incidents in these two locations. Perhaps a broader approach, examining demographic and contextual factors outside the NYC shooting dataset, could provide more insights.
```{r plot_incident_by_Location}

# Aggregate the number of incidents per borough
incidents_by_location <- NY_shooting_cases_Updated %>%
  group_by(BORO) %>%
  summarise(total_incidents = n())

# Plot number of incidents by location
ggplot(incidents_by_location, aes(x = BORO, y = total_incidents, fill = BORO)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Incidents in Each Location", x = "Location", y = "Number of Incidents") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3")



```

### Number of Precincts per Location
An interesting observation is that Manhattan, which has the second lowest shooting incidents, has about the same number of precincts as Brooklyn, which has significantly higher shooting incidents. Conversely, the Bronx, with the second highest shooting incidents, has significantly fewer precincts than Manhattan. This raises questions about whether precincts are optimally allocated across locations. Further analysis is necessary to understand whether the current distribution of precincts effectively addresses the unique needs and challenges of each borough.
```{r number_of_precincts}
# Calculate number of unique precincts in each Location
precincts_by_location <- NY_shooting_cases_Updated %>%
  group_by(BORO) %>%
  summarise(precinct_count = n_distinct(PRECINCT))

# Plot number of precincts in each Location
ggplot(precincts_by_location, aes(x = BORO, y = precinct_count, fill = BORO)) +
  geom_bar(stat = "identity") +
  labs(title = "Number of Precincts in Each Location", x = "Location", y = "Number of Precincts") +
  theme_minimal() +
  scale_fill_brewer(palette = "Set3")
```

### Incidents by Year and Location
The stacked chart which depicts incidents by year and location, shows a consistent ratio of shooting incidents each year across different locations. This consistency suggests that the factors shooting incidents might be factors in common to all location. 
```{r incidents_by_year_boro}
# Calculate incidents by year and location
incidents_by_year_boro <- NY_shooting_cases_Updated %>%
  group_by(OCCUR_YEAR, BORO) %>%
  summarise(count = n()) %>%
  ungroup()

# Plot stacked bar chart of incidents by year and location
ggplot(incidents_by_year_boro, aes(x = OCCUR_YEAR, y = count, fill = BORO)) +
  geom_bar(stat = "identity") +
  labs(
    title = "Incidents by Year and Location",
    x = "Year",
    y = "Number of Incidents",
    fill = "Location"
  ) +
  theme_minimal() +
  theme(legend.position = "right") +
  scale_fill_brewer(palette = "Set3")


```

### Incidents Over Time and the Precinct Density
The red line chart reveals a notable trend in shooting incidents over time, particularly showing a steady decline post-2011, with the lowest recorded incidents occurring between 2017 and 2019, approximately half compared to the pre-2011 levels. However, after 2019, there was a significant increase in the shooting incidents observed prior to 2011. Meanwhile, the number of precincts did not significantly increase or decrease over the years. Despite this stability in precinct density, the fluctuations in shooting incidents indicate that factors beyond the presence of precincts are influencing these trends. 
```{r plot_incident_over_time}
# Calculate total incidents per year
incidents_per_year <- NY_shooting_cases_Updated %>%
  group_by(OCCUR_YEAR) %>%
  summarise(total_incidents = n())

# Calculate total precincts per year
precincts_per_year <- NY_shooting_cases_Updated %>%
  group_by(OCCUR_YEAR) %>%
  summarise(total_precincts = n_distinct(PRECINCT))

# Calculate max values for scaling
incidents_max <- max(incidents_per_year$total_incidents)
precincts_max <- max(precincts_per_year$total_precincts)

# Merge datasets
combined_data <- merge(incidents_per_year, precincts_per_year, by = "OCCUR_YEAR", all = TRUE)

# Normalize precincts data relative to incidents scale using combined_data
combined_data$normalized_precincts <- combined_data$total_precincts / precincts_max * incidents_max

# Plot overlay of total incidents and normalized precincts per year
ggplot(combined_data, aes(x = OCCUR_YEAR)) +
  geom_line(aes(y = total_incidents, color = "Incidents"), size = 1.2) +
  geom_line(aes(y = normalized_precincts, color = "Precincts"), size = 1.2) +
  scale_y_continuous(name = "Total Incidents",
                     sec.axis = sec_axis(~ . * precincts_max / incidents_max,
                                         name = "Total Precincts",
                                         breaks = seq(0, precincts_max, by = 10))) +
  labs(title = "Total Incidents vs Total Precincts per Year",
       x = "Year",
       color = "Legend") +
  scale_color_manual(values = c("Incidents" = "red", "Precincts" = "blue")) +
  theme_minimal()
```

### Top 3 Years with Lowest Shooting Incidents
```{r find_top3_lowest_incident_in_time}
# Selecting relevant columns
shooting_data <- NY_shooting_cases_Updated %>%
  select(OCCUR_YEAR)

# Counting incidents per year
incident_counts <- table(shooting_data$OCCUR_YEAR)

# Converting incident counts to data frame
incident_counts_df <- as.data.frame(incident_counts)

# Renaming columns for clarity
colnames(incident_counts_df) <- c("Year", "Incident_Count")

# Sorting incidents in ascending order
sorted_incidents <- incident_counts_df[order(incident_counts_df$Incident_Count), ]

# Selecting top 3 years with lowest shooting incidents
top_3_years <- head(sorted_incidents, 3)

# Displaying the top 3 years and their incident counts
top_3_years
```


## Statistical Analysis
### Linear Regression Model
The analysis found that about 34.48% of the variability in shooting incidents can be explained by the number of precincts. The p-value for precinct count is not significant, indicating that the relationship between precinct count and shootings is weak based on the data. Other factors not considered in the model might affect shooting incidents.


```{r linear_regression_model}
# Simple linear regression: predicting number of incidents by precinct 
# in each location
# Combine the two datasets: 
# the number of incidents per location and 
# total number of precincts in each location
combined_data2 <- incidents_by_location %>%
  inner_join(precincts_by_location, by = "BORO")

# Fit a linear regression model
model <- lm(total_incidents ~ precinct_count, data = combined_data2)

# Display the summary of the model
summary(model)
```

## Bias and Validity of Analysis
### Identification of Bias
The interpretation of the analysis of shooting incidents may be influenced by media coverage of police brutality or biases regarding excessive use of force by law enforcement. These external influences can shape the analyst's perception and potentially skew the analysis.

### Validity of Analysis
Linear regression analysis is purely data-driven and was used to reduce biases in estimating statistical relationships between precinct count and shooting incidents. The report identifies the lack of a significant relationship within the dataset between the number of precincts per location and the number of shooting incidents in the same location.


## Conclusion
The statistical analysis in this report does not support the impact of the number of precincts on reducing shooting incidents. This suggests that other factors beyond precinct count likely influence the number of shooting incidents. To understand the root causes of the shooting incidents, we could further analyze demographic variables in the dataset, such as the age group, sex, and race of both perpetrators and victims. These demographic variables may provide insights associated with shooting incidents, thus enabling the development of effective interventions. 


## References
1. NYPD Shooting Incident Data (Historic): https://catalog.data.gov/dataset/nypd-shooting-incident-data-historic
2. R Plotting: https://www.w3schools.com/r/r_graph_plot.asp
3. R Plot Function: https://www.datamentor.io/r-programming/plot-function
